The Warrior Stabs the Goat: Creating a Dothraki Resource Grammar with Grammatical Framework

### 2016-10-14 documentation/paper.tex
More discussion in doc

\subsection{Limitations}
To begin programming a resource grammar for the first time was 
not easy, since it is not possible to start with, for example, 
just one category and then test the implementation. As there are 
so many dependencies in the resource grammar library, it is recommended 
to take the resource grammar of a similar language and change it step 
by step. But this way, as long as our implementation of the Dothraki 
language was incomplete, the resource grammar would fill the gaps with 
the wrong language grammar. Another way is to start out with an 
independent mini resource grammar, but in that case you cannot make 
use of the resources of the library. Finding our own solution for
this problem took time, but in the end we are satisfied with how
it turned out.

For the implementation of a resource grammar, a good grammar
source is needed (Grammatical Framework, 2016b). Although the
full resource grammar for the Dothraki language is not yet complete,
it is becoming harder and harder to find reliable resources regarding
further aspects of grammar we would like to implement (e.g.
usage of structural words). As it is a constructed language, and
relatively new, we hope that further and more precise information
will be documented and published with time. 
Yet, even if we had all the information needed for a full implementation,
it would not have been possible for us to write a full resource grammar
in such a short amount of time, especially as we are still new to the
GF programming language and not experienced when it comes to implementing
resource grammars. Hence, we focus on providing a good basis of implemented
rules, so that we – and others – can expand the project in the future.

>>>>

\subsection{Difficulties}
Beginning to write a resource grammar for the first time was 
not easy. Due to the structure of the full RGL, it is not 
possible to implement a resource grammar for a new language
step-by-step, adding linearizations for a few important
categories and expanding gradually, at least not when basing
it directly on the abstract specifications from the full RGL.

Of course, it is not feasible either to implement the entire
resource grammar before being able to do any kind of testing.
As a remedy for this problem, Ranta recommends
taking the resource grammar of a similar language as a
starting point and changing it step 
by step \cite[p.\ 224]{gf-book}. 
This proved difficult in our case, since Dothraki, as a constructed
language, of course does not have family ties to any existing language. 
While its grammar is certainly inspired by phenomena seen in
natural languages, these appear to be picked from a variety
of different language families (many of which, as non-linguists, we are not
particularly familiar with). Furthermore, taking
a language like English as a starting point, which is certainly similar
to Dothraki in many aspects, but wildly different in others,
proved to be fraught with difficulties as well. This is because
of the web of dependencies in a grammar,
changes at one point, such as increasing the number of cases 
from two to five, tend to require changes in a number
of different places all over the grammar, an effect that
can quickly snowball. Finding our own solution for
this problem, namely growing the abstract part gradually 
(by selectively copying over parts from the full RGL) along with
the concrete implementation, took time, but in the end we 
are satisfied with how it turned out.

\subsection{Limitations}
As Ranta notes in \cite{gf-slides}, the most important
sources for creating a resource grammar are
``[a] \emph{good} grammar book'' and ``[a] \emph{good} dictionary''.

That poses some problems in the case of Dothraki.
There are basically only two sources of information
on Dothraki grammar: The writings of inventor 
David J.\ Peterson \cite{peterson-web},
the community-operated ``Tongues of Ice and Fire`` wiki
at \url{wiki.dothraki.org}. These tend to be great resources
for the basic parts of the language. Peterson's
blog contains detailed, accessible explanations of
many grammatical phenomena featured in
the snippets of Dothraki, that have appeared on the show
up to its third season. And the enthusiastic Dothraki community
has done great work organizing and detailing this information.

However, as our work progressed, it has become harder and harder
to find good information on some of the more subtle or
obscure parts of the grammar. It should be noted that the
total corpus of text in the Dothraki language 
still comprises only a few hundred sentences.
By now, we have already mostly
incorporated the information available from the two sources
mentioned above (the two largest remaining omissions are wh-questions and
numerals). In many cases we have run up against the limits
of what information is available and sometimes we have been
forced to resort to the Dothraki corpus itself
in an effort to try and reverse engineer
language features that so far have not been explained by
their inventor.

We give three examples:
\begin{itemize}
  \item As noted in Section \ref{sec-verbs} our resource grammar
  so far only contains a dummy implementation of
  verbs with a verb phrase complement (§VV§s). In English, 
  (auxiliary) verbs like ``can'', ``must'' and ``want to'' are
  of this kind. In Dothraki, as explained in the quoted section,
  ``can'' and ``must'' are rendered as particles
  (``laz'' and ``eth'') and this is what our current
  implementation captures. However, certainly not all
  §VV§s are rendered like this
  in Dothraki. These non-inflected particles almost surely
  are a small closed class of words, whereas §VV§s are
  plentiful (think ``agree to (doing sth)'', ``consider (doing sth)'',
  ``enjoy (doing sth)'', ``expect to (do sth)'' etc.).
  However, there is almost no explanation available
  on the syntactical structure of such verbs in Dothraki.

  When they have appeared on the show, verb phrase
  complements are usually avoided in favour of
  embedded clauses:

  \begin{quote}
  ``The khaleesi wants to eat something different tonight.'' =
  ``Khaleesi zala meme adakha esinakh ajjalan.'' (lit.\ ``The khaleesi wants that she eats something different tonight'')\\
  ``You'd rather be sold into slavery?'' = ``Yer zali meyer nem vazhi ven zafra?'' (lit.\ ``You want that you are sold into slavery?'')
  \end{quote}

  As far as we can tell, in the entire corpus there are only four sentences
  in which the verb ``zalat'' (to want) is used with what seems to be a
  true verb phrase complement:
  \begin{quote}
    ``Hash shafka zali addrivat mae [...]?'' = ``Do you want me to kill him?''\\
    ``[...] hash yer zali zifichelat moon?'' = ``[...] you want to steal from him?''\\
    ``Anha zalak nesat ven fini ven athyazhar khaleesisi.'' = ``I'd like to know what a Khaleesi tastes like.''\\
    ``Hash yeri vo zali nesat rek dirgak anha?'' = ``Don't you want to know what I think?''
  \end{quote}
  We could not find any other instances of §VV§ verbs being used
  with a true verb phrase complement in the corpus.

  From these four sentences it certainly appears like verb phrase complements
  (similar to English) are formed with the verb in the infinitive
  (``addrivat'', ``zifichelat'' and ``nesat'' are all infinitive forms).
  However, as far as we can tell, this has not been discussed by 
  Peterson so far, nor is there any information about it on the wiki.

  It seems rather premature and speculative
  to conclude from just four sentences (all featuring
  the same verb) that this is the general way verb phrase
  complements are rendered in Dothraki. Therefore, we are at an
  impasse: Until further information becomes available,
  we cannot implement §VV§s (other than the few
  special ones like ``can'' and ``must'').

  \item The module \name{Structural} contains a number
  of common structural words that should be available in 
  every language. Quite a few of these are simply not yet
  available in Dothraki (``somewhere'', ``almost'',
  ``quite'', ``otherwise'', ``at least'' etc.), but
  even those that have translations in the vocabulary list
  are sometimes difficult to implement. For example, the
  vocabulary list notes that ``someone'' translates as
  ``ato''. But since this is a noun phrase, it probably
  must inflect somehow.
  Inflection in nouns depends on their animacy, but
  it is not clear whether inanimate or animate declension
  is used here (or some irregular inflection, as with determiners). 
  In the corpus ``ato'' appears exactly once,
  in the nominative. In the end, we had no other choice
  but to guess the declension (we decided on standard
  animate inflection).

  Similarly, ``eyak'' supposedly means ``everybody'', according
  to the vocabulary, but in the only place it is used
  in the corpus, it seems to be used in the sense of ``everything''.
  In that instance it is used in the genitive (probably; the phrase
  is ``hatif eyaki'' and the preposition ``hatif'' can assign
  genitive, allative or ablative case, depending on meaning).
  In the end we decided on a declension based on this one occurence and
  by analogy with ``vosak'' (= ``nobody''), which occurs in
  the accusative once.

  In the end, in many cases in the \name{Structural} module, 
  we had to employ something more akin to educated guesswork rather
  than being able to rely on definitive sources.

  \item §Comp§s (``complementizers to a copula'') can be used
  to form imperatives: ``Be strong!'' or ``Be a warrior!''.
  The first has an obvious rendering in Dothraki (by putting
  the verb ``hajat'' = ``to be strong'' derived from ``haj'' = ``strong''
  in the imperative), but the second does not. In Dothraki
  there is no analogue of the copula ``to be'' that could be
  used in the imperative here. In the corpus, no sentence
  like this exists. For now, we cannot implement this
  correctly.
\end{itemize}

We are hopeful that, in spite of these current limitations,
our work so far can serve as a good basis for implementing
more language features as further and more precise information 
on them becomes available.
